A Large-Scale Study of the Use of Eval in R

### 2020-08-28 inst/paper/main.tex
Minor

Table~\ref{table:site-package-summary} shows the number of call sites in
CRAN packages. 61 packages have a single call site, and one package, \emph{VGAM}, has \MaxCallSitePackage
call sites of \eval. \NoCallSitePackages packages have no calls to \eval. The core packages
have \NbCoreEvalCallSites call sites of \eval and Kaggle has none. These numbers are
under approximations as \eval may be called reflectively, through aliases,
or passed to a higher-order function.

@@

To roughly estimate the complexity of the code being evaluated,
Fig.~\ref{fig:ast-size} shows the distribution of the abstract syntax tree
size of the \c{expr} argument to \eval. The majority of inputs across core
and packages (\NbAstOnePercent of \eval calls and \NbAstOneCallSitePercent
of call sites) have size 1. That is to say, the code being evaluated
consists of a single variable name or constant. A function call needs at
least three nodes and a list such as \c{c(1, 2, 3)} counts for 4
nodes. Inputs smaller than \AstSizeNineFive nodes account for 95\% of all
calls. Sizes in core are larger than in packages. Most large AST originate
from library \c{VGAM}, which is used for statistical modeling, and also calls functions using \eval from core. It uses
\eval to evaluate large matrices with values in a given environment.

@@

We observe a very wide spread in the number of events generated by \eval
calls. The largest number of events generated by an \eval call is
\EventsMaxCount.  However majority of \eval calls perform up to
\EventsMaxRangeA events suggesting that most expressions passed to \eval are
small.  It is interesting to note that a higher proportion of Core R \eval
calls perform side effects in the \EventsMinRangeB--\EventsMaxRangeB range.
Only \EventsCoreEvalCountC Core R \eval calls and \EventsPackageEvalCountC
package \eval call generate \EventsMinRangeC to \EventsMaxRangeC events.
These \eval calls originate from statistical modeling packages such as
\mlogit, \mboost, \metafor, \lavaan, \mclust and \gamlss.

\medskip\noindent\emph{Discussion:} Comparing the use of \eval in R and
JavaScript is instructive. The use of \eval is more widespread in data
science code than it was in JavaScript. Any run of an R program seems to
trigger \eval, more over package programmers use \eval liberally. On the
other hand, the authors of Kaggle packages do not use the feature at all.
This point to a bifurcated user community, experts who write package know
how to use \eval while end-users do not. An alternative explanation is that
end-users do not need \eval for simple data analysis tasks. This bifurcation
was not observed in the JavaScript community, but this may have changed with
the advent of Node.js ecosystem.  The distribution of input sizes is similar
to that found in JavaScript, the majority of inputs being rather small.

>>>>

Table~\ref{table:site-package-summary} shows the number of call sites in
CRAN packages. 61 packages have a single call site, and one package, \emph{VGAM}, has \MaxCallSitePackage
call sites of \eval. \NoCallSitePackages packages have no calls to \eval. The core packages
have \NbCoreEvalCallSites call sites of \eval and Kaggle has none. These numbers are
under-approximations as \eval may be called reflectively, through aliases,
or passed to a higher-order function.

@@

To roughly estimate the complexity of the code being evaluated,
Fig.~\ref{fig:ast-size} shows the distribution of the abstract syntax tree
size of the \c{expr} argument to \eval. The majority of inputs across core
and packages (\NbAstOnePercent of \eval calls and \NbAstOneCallSitePercent
of call sites) have size 1. That is to say, the code being evaluated
consists of a single variable name or constant. A function call needs at
least three nodes and a list such as \c{c(1, 2, 3)} counts for 4
nodes. Inputs smaller than \AstSizeNineFive nodes account for 95\% of all
calls. Sizes in core are larger than in packages. Most large ASTs originate
from library \c{VGAM}, which is used for statistical modeling, and also calls functions using \eval from core. It uses
\eval to evaluate large matrices with values in a given environment.

@@

We observe a very wide spread in the number of events generated by \eval
calls. The largest number of events generated by an \eval call is
\EventsMaxCount.  However, a majority of \eval calls performs up to
\EventsMaxRangeA events suggesting that most expressions passed to \eval are
small.  It is interesting to note that a higher proportion of Core R \eval
calls perform side effects in the \EventsMinRangeB--\EventsMaxRangeB range.
Only \EventsCoreEvalCountC Core R \eval calls and \EventsPackageEvalCountC
package \eval call generate \EventsMinRangeC to \EventsMaxRangeC events.
These \eval calls originate from statistical modeling packages such as
\mlogit, \mboost, \metafor, \lavaan, \mclust and \gamlss.

\medskip\noindent\emph{Discussion:} Comparing the use of \eval in R and
JavaScript is instructive. The use of \eval is more widespread in data
science code than it was in JavaScript. Any run of an R program seems to
trigger \eval; moreover package programmers use \eval liberally. On the
other hand, the authors of Kaggle packages do not use the feature at all.
This point to a bifurcated user community: experts who write package know
how to use \eval while end-users do not. An alternative explanation is that
end-users do not need \eval for simple data analysis tasks. This bifurcation
was not observed in the JavaScript community, but this may have changed with
the advent of Node.js ecosystem.  The distribution of input sizes is similar
to that found in JavaScript, the majority of inputs being rather small.
